Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-13 submission/manuscript.Rmd
First pass at Results

@@

## Abstract

@@

## Introduction (400 words)

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Haggar2009; @Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental,  behavioral, and other factors such as smoking and diet [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as mutated cells proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis [@Flynn2016]. Numerous studies in murine models have supported this model [@Zackular2013; @Baxter2014; @Zackular2015; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Kostic2011; @Sze2018; @Baxter2016; @Zackular2014; @Zeller2014; @Shah2017; @Dejea2018; @Arthur2012]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

@@

To better understand the connection between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of SCFAs in feces of previously characterized individuals with normal colons, adenomas, and carcinomas and subset of those individuals after they underwent treatment for their lesions. Our goals included: (1) testing whether there was an association between SCFA concentration and tumor status, (2) determining whether SCFA concentrations could be used as biomarkers to improve the detection of colonic lesions, and (3) predicting SCFA concentrations based on the relative abundance of fecal bacteria and their genes.

## Results (700 words)

* There is no association between presence of tumors and SCFA concentration
	- Wilcoxon/T-test
	- Cross sectional
	- Aggregating SCFAs does not improve association with CRC
	- Combining 16S and SCFA does not improve prediction of CRC
	- Lack of change pre and post treatment
* Community structure is not predictive of SCFA production
	- Random forest regression
	- 16S
	- picrust
	- shotgun metagenomic data

## Discussion (200 words)

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @colorectal cancer_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =`r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())` and carcinoma n = `r length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @colorectal cancer_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@scfa_measures_venkataraman2016]. The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the actual weight of fecal samples instead of the average weight for SCFA concentration normalizations. These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@scfa_measures_venkataraman2016] and Figure 1 here).

**16S rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @normalization_sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@opticlust_westcott2017].

@@

**Obtaining Operational Protein Families from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()`, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017]. Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015]. Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF alignment used the KEGG database [@Kanehisa2015].

**Pulling genes involved with SCFA synthesis.** Specific genes located near the end of the pathways involved in the synthesis of acetate, butyrate, and propionate were analyzed for any differences between individuals with normal colons and those with tumors. These genes were based on pathways from KEGG as well as previous research [@scfa_baxter2014; @Kanehisa2015] and a list can be found in the supplemental material [Table S1].

**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification and regression models via the caret package [@randomforest_citation_2002; @caret_citation]. This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations. The regression models were built to classify the SCFA concentrations of acetate, butyrate, and propionate regardless of disease status.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Wallis rank sum test with a Dunn's post-hoc test was used to assess differences between individuals without colon tumors, patients with adenomas, and patients with carcinomas. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, the imputed gene counts of genes encoding enzymes involved in SCFA synthesis was tested. Additionally, metagenomic sequencing counts for important genes involved with SCFA production were analyzed. From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without colon tumors and patients with adenomas or carcinomas using Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models. Finally, models to classify the actual SCFA concentration or high/low SCFA concentration based on the median of each SCFA using 16S rRNA gene sequencing data was created using the Random Forest algorithm. For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.

@@

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chain fatty acid analysis on the High-Performance Liquid Chromatography machine at the University of Michigan. We would also like to thank Ada Hagan for providing valuable feedback on earlier drafts of the manuscript. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure 1. No differences in SCFA measurements were observed between individuals with normal colons or those with adenoma or carcinomas before treatment (A) or due to treatment (B).** Concentrations of SCFAs did not differ among individuals in the different diagnosis categories (normal=XXX, adenoma=XXX, carcinoma=XXX) except for isobutyrate, which was significantly higher among individuals with cancer than those with adenomas (P=0.XXXX), but no different from those with normal colons (all P>0.XXX) (A). The change in SCFA concentrations among individuals undergoing treatment who initially had adenoma (N=XX) and carcinoma (N=XX) was only significant for the isobutyrate concentration among individuals with carcinoma (P=0.XXX, all others P>0.XXX).

**Figure 2. SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions.**

**Figure 3. 16S rRNA gene and metagenomic sequence data do not predict SCFAs concentrations.**

>>>>

```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
options(tidyverse.quiet = TRUE)

library(tidyverse)
library(knitr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
  print(x)
  if(is.list(x)){
    x <- unlist(x)
  }
  if(is.numeric(x)){
    if(abs(x - round(x)) < .Machine$double.eps^0.5){
      paste(format(x,big.mark=',', digits=0, scientific=FALSE))
    } else {
      paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
    }
  } else {
      paste(x)
  }
}
knitr::knit_hooks$set(inline=inline_hook)
```

@@

## Abstract (250 words)

Something clever

## Importance (150 words)

Something clever

@@

Colorectal cancer is the third leading cancer-related cause of death within the United States [@Haggar2009; @Siegel2016]. Less than 10% of cases can be attributed to genetic risk factors [@Fearon1990]. This leaves a significant role for environmental,  behavioral, and other factors such as smoking and diet [@FlissIsakov2017; @Lee2015]. Colorectal cancer is thought to be initiated by a series of mutations that accumulate as mutated cells proliferate leading to adenomatous lesions, which are succeeded by carcinomas [@Fearon1990]. Throughout this progression, there are ample opportunities for bacterial populations to create mutations, induce inflammation, and accelerate tumorigenesis [@Flynn2016]. Numerous studies in murine models have supported this model [@Zackular2013; @Baxter2014; @Zackular2015; @DeStefanoShields2016; @Tomkovich2017]. Additional cross sectional studies in humans have identified microbiome-based biomarkers of disease [@Kostic2011; @Sze2018; @Baxter2016; @Zackular2014; @Hannigan2017; @Zeller2014; @Shah2017; @Dejea2018; @Arthur2012]. These studies suggest that in some cases, it is the loss of bacterial populations that produce short-chain fatty acids (SCFAs) that results in increased inflammation and tumorigenesis.

@@

```{r, concentration}
n <- read_csv('../data/metadata/cross_section.csv') %>% count(dx)
n_normal <- n %>% filter(dx == "normal") %>% pull(n)
n_adenoma <- n %>% filter(dx == "adenoma") %>% pull(n)
n_cancer <- n %>% filter(dx == "cancer") %>% pull(n)

cross_test <- read_tsv('../data/process/scfa_cross_section_stats.tsv')
iso_p_cross <- cross_test %>% filter(p.value < 0.05)
stopifnot(nrow(iso_p_cross) == 1)

n_follow <- read_csv('../data/metadata/follow_up.csv') %>% count(dx)
n_follow_adenoma <- n_follow %>% filter(dx == "adenoma") %>% pull(n)
n_follow_cancer <- n_follow %>% filter(dx == "cancer") %>% pull(n)

follow_test <- read_tsv('../data/process/scfa_pre_post_stats.tsv')
iso_p_follow <- follow_test %>% filter(p.value < 0.05)
stopifnot(nrow(iso_p_follow) == 1)

pooled_test <- min(c(cross_test %>% filter(scfa=="pooled") %>% pull(p.value), follow_test %>% filter(scfa=="pooled") %>% pull(p.value)))

```

**SCFA concentrations do not meaningfully vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, isobutyrate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016]. The only SCFA that had a significantly different concentration across the diagnoses was isobutyrate (P=`r iso_p_cross$p.value`; Figure 1A). The median concentration of isobutyrate was `r iso_p_cross$normal` mmol/kg in people with normal colons and it was `r iso_p_cross$adenoma` and `r iso_p_cross$cancer` mmol/kg in people with colonic adenomas or carcinomas. The difference in isobutyrate concentration between people with adenomas and carcinomas was significantly different (P=`r iso_p_cross["cancer-adenoma"]`); however, the differences between people with normal colons and those with adenomas or carcinomas was not significant (P=`r iso_p_cross["normal-adenoma"]` and P=`r iso_p_cross["normal-cancer"]`). The median concentration of isobutyrate in people with normal colons (`r iso_p_cross$normal` mmol/kg) was between those with adenomas (`r iso_p_cross$adenoma` mmol/kg) and carcinomas (`r iso_p_cross$cancer` mmol/kg). Among the subjects with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017]. The only SCFA that changed following treatment was isobutyrate, which decreased by `r iso_p_follow$estimate` mmol/kg (P=`r iso_p_follow$p.value`; Figure 1B). For both the pre-treatment cross-sectional data and the pre/post treatment data, we pooled the SCFA concentrations on a per molecule of carbon basis and again failed to see any significant differences (P>`r pooled_test`). The low concentration of isobutyrate relative to the other SCFAs, inconsistent concentrations, and unexpected decrease in concentration with treatment makes it difficult to ascribe much biological relevance to this observation.

```{r, classification}
```

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models. We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model. A model trained on the individual or pool SCFAs alone had a median area under the receiver operator characteristic curve (AUROC) of **0.XX** (IQR=**0.XX**). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances (AUROC~OTU+SCFA~=**0.XX**, AUROC~genus+SCFA~=**0.XX**), the AUROC values were no different from the models trained without the SCFA concentrations (AUROC~OTU~=**0.XX**, AUROC~genus~=**0.XX**; Figure 2A). We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=**XX**, N~adenoma~=**XX**, and N~cancer~=**XX**). We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories [**XXXXXX**]. Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.

```{r, regression}
```

**Knowledge of microbial community structure does not predict SCFA concentrations.** Regardless of a person's diagnosis, we next asked whether the fecal community structure was predictive of fecal SCFA concentrations. We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the individual and pooled SCFAs. Regardless of the binning method or SCFA, the largest amount of variation that the model could explain was **0.XXXX** (Figure 2B). Next, we trained Random Forest regression models using metagenomic sequence data binned into OPFs and KEGG categories to predict the concentration of the individual and pooled SCFAs. Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways, enzyme commission numbers, or KEGG categories using picrust2. Again, SCFA concentrations could not be predicted based on the imputed metagenomic data; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and genes is insufficient to predict SCFA concentrations.

**Conclusion.**

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@Sze2017; @Baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=**r length(metai$sample)**) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =**r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())** and carcinoma n = ** length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())**). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from **r min(metaf$time)** - **r max(metaf$time)** days. Our use of treatment has been previously defined as encompassing removal of a tumor (surgery or colonoscopy) with or without chemotherapy and radiation [@Sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@Sze2017; @Baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Declaration.

**Measuring specific SCFAs.** Our protocol for the measurement of acetate, butyrate, and propionate followed a previously published protocol that used a High-Performance Liquid Chromatography (HPLC) machine [@Venkataraman2016]. The following changes to this protocol included the use of frozen fecal samples suspended in 1ml of PBS instead of fecal suspensions in DNA Genotek OmniGut tubes, and the use of the actual weight of fecal samples instead of the average weight for SCFA concentration normalizations. These methodological changes did not affect the overall median concentrations of these SCFAs between the two studies (see Table 1 [@Venkataraman2016] and Figure 1 here).

**16S rRNA gene sequencing.** The workflow and processing have been previously described [@Schloss2009; @Kozich2013; @Sze2017]. In brief, sequences were quality filtered and contigs created from the paired end reads. Any sequences with ambiguous base calls were discarded. Contigs were then checked for matches to the V4 region of the 16S rRNA gene using the SILVA database [@Quast2012]. Chimeras were identified and removed using UCHIME and OTUs clustered at 97% similarity [@Edgar2011]. The major differences from these previous reports include: the use of version 1.39.5 of the mothur software package and clustering Operational Taxonomic Units (OTUs) at 97% similarity using the OptiClust algorithm [@Westcott2017].

@@

**Obtaining Operational Protein Families from metagenomes.** A subset of the cross-sectional group (n=**r length(metai$sample)**) containing a total of **r length(geof_samples$X1)** individuals (normal n=**r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()**, adenoma n=**r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()**, and carcinoma n=**r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()**) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017]. Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015]. Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF alignment used the KEGG database [@Kanehisa2015].

**Random Forest models.** The model was first trained on 80% of the data and then tested on the held out 20% (80/20 split) using the Random Forest algorithm for classification and regression models via the caret package [@Liaw2002; @Kuhn2017]. This was repeated on 100 different 80/20 splits of the data to generate a reasonable range for the AUC of the model. The reported AUCs, unless otherwise specified, are for the test sets. The classification models were built to group normal versus adenoma, normal versus carcinoma, and high versus low SCFA concentrations. The regression models were built to classify the SCFA concentrations of acetate, butyrate, and propionate regardless of disease status.

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, a Kruskal-Wallis rank sum test with a Dunn's post-hoc test was used to assess differences between individuals without colon tumors, patients with adenomas, and patients with carcinomas. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [**XXXXXX**]. First, we assessed differences in SCFA concentrations measured by HPLC between individuals with normal colons and patients with tumors (adenoma or carcinoma). We then analyzed whether SCFA concentrations changed in patients with an adenoma or carcinoma pre- versus post-treatment. Next, the imputed gene counts of genes encoding enzymes involved in SCFA synthesis was tested. Additionally, metagenomic sequencing counts for important genes involved with SCFA production were analyzed. From here we analyzed the number of significant positive and negative correlations between OTU relative abundance and SCFA concentrations in individuals without colon tumors and patients with adenomas or carcinomas using Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs were better able to classify individuals with and without tumors using Random Forest models. Finally, models to classify the actual SCFA concentration or high/low SCFA concentration based on the median of each SCFA using 16S rRNA gene sequencing data was created using the Random Forest algorithm. For all Random Forest models, the assessment of the most important variables was based on the top 10 features (OTUs or SCFAs) using the mean decrease in accuracy.

@@

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would thank the University of Michigan Center for Microbial Systems for enabling our short-chain fatty acid analysis. Support for MAS came from the Canadian Institute of Health Research and the National Institutes of Health (UL1TR002240). Support for PDS came from the National Institutes of Health (P30DK034933 and R01CA215574).

@@

\includegraphics{../results/figures/scfa_abundance.pdf}

**Figure 1. SCFA concentrations did not vary meaningfully with diagnosis of colonic lesions or with treatment for adenomas or carcinomas.** (A) We measured the concentration of fecal SCFAs from individuals with normal colons (N=`r n_normal`) or those with adenoma (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) was for isobutyrate. (B) A subset of individuals diagnosed with adenomas (N=`r n_follow_adenoma`) or carcinomas (N=`r n_follow_cancer`) who underwent treatment were resampled a year after the initial sampling; one extreme propionate value (124.4 mmol/kg) was included in the adenoma analysis but censored from the visualization for clarity.

**Figure 2. SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions. 16S rRNA gene and metagenomic sequence data do not predict SCFAs concentrations.**

### 2019-03-13 submission/manuscript.tex
First pass at Results

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

@@

\hypertarget{introduction-400-words}{%
\subsection{Introduction (400 words)}\label{introduction-400-words}}

@@

To better understand the connection between colorectal cancer, the
microbiome, and SCFAs, we quantified the concentration of SCFAs in feces
of previously characterized individuals with normal colons, adenomas,
and carcinomas and subset of those individuals after they underwent
treatment for their lesions (14, 16). Our goals included: (1) testing
whether there was an association between SCFA concentration and tumor
status, (2) determining whether SCFA concentrations could be used as
biomarkers to improve the detection of colonic lesions, and (3)
predicting SCFA concentrations based on the relative abundance of fecal
bacteria and their genes.

\hypertarget{results-700-words}{%
\subsection{Results (700 words)}\label{results-700-words}}

\begin{itemize}
\tightlist
\item
  There is no association between presence of tumors and SCFA
  concentration

  \begin{itemize}
  \tightlist
  \item
    Cross sectional
  \item
    Aggregating SCFAs does not improve association with CRC
  \item
    Lack of change pre and post treatment
  \item
    Aggregating SCFAs does not improve association with CRC
  \end{itemize}
\item
  Adding SCFA data to diagnostic models does not improve performance

  \begin{itemize}
  \tightlist
  \item
    Combining 16S and SCFA does not improve prediction of CRC
  \item
    OTU, genus, OPF, kegg
  \end{itemize}
\item
  Community structure is not predictive of SCFA production

  \begin{itemize}
  \tightlist
  \item
    Random forest regression
  \item
    16S
  \item
    picrust
  \item
    shotgun metagenomic data
  \end{itemize}
\end{itemize}

\hypertarget{discussion-200-words}{%
\subsection{Discussion (200 words)}\label{discussion-200-words}}

@@

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (8,
49) and a list can be found in the supplemental material {[}Table S1{]}.

@@

\textbf{Figure 1. No differences in SCFA measurements were observed
between individuals with normal colons or those with adenoma or
carcinomas before treatment (A) or due to treatment (B).} Concentrations
of SCFAs did not differ among individuals in the different diagnosis
categories (normal=\textbf{XXX}, adenoma=\textbf{XXX},
carcinoma=\textbf{XXX}) except for isobutyrate, which was significantly
higher among individuals with cancer than those with adenomas
(P=\textbf{0.XXXX}), but no different from those with normal colons (all
P\textgreater{}\textbf{0.XXX}) (A). The change in SCFA concentrations
among individuals undergoing treatment who initially had adenoma
(N=\textbf{XX}) and carcinoma (N=\textbf{XX}) was only significant for
the isobutyrate concentration among individuals with carcinoma
(P=\textbf{0.XXX}, all others P\textgreater{}\textbf{0.XXX}).

\textbf{Figure 2. SCFA concentrations do not improve models for
diagnosing the presence of adenomas, carcinomas, or all lesions.}

\textbf{Figure 3. 16S rRNA gene and metagenomic sequence data do not
predict SCFAs concentrations.}

>>>>

\hypertarget{abstract-250-words}{%
\subsection{Abstract (250 words)}\label{abstract-250-words}}

Something clever

\hypertarget{importance-150-words}{%
\subsection{Importance (150 words)}\label{importance-150-words}}

Something clever

@@

@@

\textbf{SCFA concentrations do not meaningfully vary with diagnosis or
treatment.} To quantify the associations between colorectal cancer, the
microbiome, and SCFAs, we quantified the concentration of acetate,
propionate, isobutyrate, and butyrate in feces of previously
characterized individuals with normal colons (N=172) and those with
colonic adenomas (N=198) or carcinomas (N=120) (14). The only SCFA that
had a significantly different concentration across the diagnoses was
isobutyrate (P=0.0091; Figure 1A). The median concentration of
isobutyrate was 3.30 mmol/kg in people with normal colons and it was
3.00 and 3.84 mmol/kg in people with colonic adenomas or carcinomas. The
difference in isobutyrate concentration between people with adenomas and
carcinomas was significantly different (P=0.0065); however, the
differences between people with normal colons and those with adenomas or
carcinomas was not significant (P=0.19 and P=0.11). The median
concentration of isobutyrate in people with normal colons (3.30 mmol/kg)
was between those with adenomas (3.00 mmol/kg) and carcinomas (3.84
mmol/kg). Among the subjects with adenomas and carcinomas, a subset
(N\textsubscript{adenoma}=41, N\textsubscript{carcinoma}=26) were
treated and sampled a year later (37). The only SCFA that changed
following treatment was isobutyrate, which decreased by 0.99 mmol/kg
(P=0.002; Figure 1B). For both the pre-treatment cross-sectional data
and the pre/post treatment data, we pooled the SCFA concentrations on a
per molecule of carbon basis and again failed to see any significant
differences (P\textgreater{}0.15). The low concentration of isobutyrate
relative to the other SCFAs, inconsistent concentrations, and unexpected
decrease in concentration with treatment makes it difficult to ascribe
much biological relevance to this observation.

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units based on 97\% similarity or into genera enabled us to
classify individuals as having adenomas or carcinomas using Random
Forest machine learning models. We repeated that analysis but added the
concentration of the individual or pooled SCFAs as possible features to
train a model. A model trained on the individual or pool SCFAs alone had
a median area under the receiver operator characteristic curve (AUROC)
of \textbf{0.XX} (IQR=\textbf{0.XX}). When we trained the models with
the SCFAs concentrations and OTU or genus-level relative abundances
(AUROC\textsubscript{OTU+SCFA}=\textbf{0.XX},
AUROC\textsubscript{genus+SCFA}=\textbf{0.XX}), the AUROC values were no
different from the models trained without the SCFA concentrations
(AUROC\textsubscript{OTU}=\textbf{0.XX},
AUROC\textsubscript{genus}=\textbf{0.XX}; Figure 2A). We also trained
models using a smaller dataset that generated shotgun metagenomic
sequencing data from a subset of our cohort
(N\textsubscript{normal}=\textbf{XX},
N\textsubscript{adenoma}=\textbf{XX}, and
N\textsubscript{cancer}=\textbf{XX}). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories {[}\textbf{XXXXXX}{]}. Again, the performance of the models
trained with the meteagenomic data did not improve when the SCFA
concentrations were added as possible features when training the model
(Figure 2A). These data demonstrate that knowledge of the SCFA profile
from a patient's fecal sample does not improve the ability to diagnose a
colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations.} Regardless of a person's diagnosis, we next asked
whether the fecal community structure was predictive of fecal SCFA
concentrations. We trained Random Forest regression models using 16S
rRNA gene sequence data binned into OTUs and genera to predict the
concentration of the individual and pooled SCFAs. Regardless of the
binning method or SCFA, the largest amount of variation that the model
could explain was \textbf{0.XXXX} (Figure 2B). Next, we trained Random
Forest regression models using metagenomic sequence data binned into
OPFs and KEGG categories to predict the concentration of the individual
and pooled SCFAs. Similar to the analysis using 16S rRNA gene sequence
data, the metagenomic data was not predictive of SCFA concentration; the
largest amount of variation that the models could explain was
\textbf{0.XXXX} (Figure 2B). Because of the limited number of samples
that we were able to generate metagenomic sequence data from, we used
our 16S rRNA gene sequence data to impute metagenomes that were binned
into metabolic pathways, enzyme commission numbers, or KEGG categories
using picrust2. Again, SCFA concentrations could not be predicted based
on the imputed metagenomic data; the largest amount of variation that
the models could explain was \textbf{0.XXXX} (Figure 2B). The inability
to model SCFA concentrations from microbiome data indicates that the
knowledge of the abundance of organisms and genes is insufficient to
predict SCFA concentrations.

\textbf{Conclusion.}

@@

@@

\includegraphics{../results/figures/scfa_abundance.pdf}

\textbf{Figure 1. SCFA concentrations did not vary meaningfully with
diagnosis of colonic lesions or with treatment for adenomas or
carcinomas.} (A) We measured the concentration of fecal SCFAs from
individuals with normal colons (N=172) or those with adenoma (N=198) or
carcinomas (N=120) was for isobutyrate. (B) A subset of individuals
diagnosed with adenomas (N=41) or carcinomas (N=26) who underwent
treatment were resampled a year after the initial sampling; one extreme
propionate value (124.4 mmol/kg) was included in the adenoma analysis
but censored from the visualization for clarity.

\textbf{Figure 2. SCFA concentrations do not improve models for
diagnosing the presence of adenomas, carcinomas, or all lesions. 16S
rRNA gene and metagenomic sequence data do not predict SCFAs
concentrations.}
